Storm surges in the Strait of Georgia simulated with a regional model

### 2015-01-05 AO_guides/manuscriptAO.tex
Addressed Keith's comment about gamma^2 for Feb 2006. Also tried to clarify use of Willmott skill score

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one. Values close to one indicate good model skill. These three statistics, along with the mean observed values and mean modelled values are reported in Table \ref{tab:statistics} for each hindcast.

@@

An additional storm surge hindcast for a Nov 16, 2006 storm surge is provided as another means of assessing model performance and is mentioned briefly here. The model performs reasonably well in this hindcast; Willmott skill scores are greater than 0.94 (Table \ref{tab:statistics}).  The best match in maximum surge occurred at Campbell River with a 6.2 cm underestimate by the model. The timing of the maximum surge at Campbell River was identical with observations and too early at the other the stations.

@@

%local pressure
Next, the surge amplitude decreases by a few centimetres at all four locations if local atmospheric pressure is not included in the surface forcing. Other findings have reported the inverse barometer effect due to atmospheric pressure contributes up to $^1/_3$ of the total surge amplitude \citep{murty1995storm}. Note that in these simulations the removal of the atmospheric pressure in the surface forcing does not entirely remove the inverese barometer effect from the model results since the sea surface height anomaly at the open boundary includes an inverse barometer component. As such, these simulations indicate the importance of the local atmospheric pressure forcing but not the inverse barometer entirely. The contribution due to the inverse barometer can be calculated with equation \ref{eq:inverse} or by the approximation that 1 hPa drop in pressure corresponds to 1 cm rise in sea level. In this event, the pressure at the Vancouver International Airport dropped to about 99 000 Pa \citep{ECClimateArchive}, corresponding to a 20 cm rise in sea level. This inverse barometer component is more inline with the findings of \citep{murty1995storm} but does not alter the conclusion that the open boundary forcing contributes most significantly. Further, the model calculates the local atmopsheric pressure forcing using a spatial gradient, and since the spatial scale of the low pressure system is large compared to the domain the local atmospheric pressure should not contribute greatly.

@@

Modelled residuals are generally too low in this hindcast (Figure \ref{fig:dec2006}), however, the timing of the surge agrees well. All Willmott skill scores rank above 0.88 and the $\gamma^2$ measure also indicates good agreement between model and observations (Table \ref{tab:statistics}). The $\gamm^2$ measure suggests that the standard deviation of the error in the modelled resdiuals is roughly half of the standard deviation of the observed residuals. Note that the performance has degraded when compared to the Feb 2006 and Nov 2006 cases. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point near by but outside of the domain) is up to 10 cm lower than the observed residual at Neah Bay, WA (a point near by but inside of the domain). Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the cause.

>>>>

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and on and values close to one indicate good model skill. Since the Willmott skill score is nondimensional, it is an approriate statistic to report when comparing skills of different variables, in this case, the total water level and the residuals. Since the total water level has a larger variability than the residuals the $\gamma^2$ measure should be much lower for the total water levels. These three statistics, along with the mean observed values and mean modelled values (which indicate the model's bias) are reported in Table \ref{tab:statistics} for each hindcast.

@@

An additional storm surge hindcast for a Nov 16, 2006 storm surge is provided as another means of assessing model performance and is mentioned briefly here. The model performs reasonably well in this hindcast; Willmott skill scores are greater than 0.94 (Table \ref{tab:statistics}).  Further, the $\gamma^2$ statistic suggests that the standard deviation of the error in the modelled residuals is about 40\% ($\sqrt{0.16}$) of the standard deviation of the observed residuals. The best match in maximum surge occurred at Campbell River with a 6.2 cm underestimate by the model. The timing of the maximum surge at Campbell River was identical with observations and too early at the other the stations.

@@

%local pressure
Next, the surge amplitude decreases by a few centimetres at all four locations if local atmospheric pressure is not included in the surface forcing. Other findings have reported the inverse barometer effect due to atmospheric pressure contributes up to $^1/_3$ of the total surge amplitude \citep{murty1995storm}. Note that in these simulations the removal of the atmospheric pressure in the surface forcing does not entirely remove the inverese barometer effect from the model results since the sea surface height anomaly at the open boundary includes an inverse barometer component. As such, these simulations indicate the importance of the local atmospheric pressure forcing but not the inverse barometer entirely. The contribution due to the inverse barometer can be calculated with equation \ref{eq:inverse} or by the approximation that 1 hPa drop in pressure corresponds to 1 cm rise in sea level. In this event, the pressure at the Vancouver International Airport dropped to about 99 000 Pa \citep{ECClimateArchive}, corresponding to a 20 cm rise in sea level. This inverse barometer contribution is more inline with the findings of \citep{murty1995storm} but does not alter the conclusion that the open boundary forcing contributes most significantly. Further, the model calculates the local atmopsheric pressure forcing using a spatial gradient, and since the spatial scale of the low pressure system is large compared to the domain the local atmospheric pressure should not contribute greatly.

@@

Modelled residuals are generally too low in this hindcast (Figure \ref{fig:dec2006}), however, the timing of the surge agrees well. All Willmott skill scores rank above 0.88 and the $\gamma^2$ measure also indicates good agreement between model and observations (Table \ref{tab:statistics}). The $\gamma^2$ measure suggests that the standard deviation of the error in the modelled resdiuals is roughly half of the standard deviation of the observed residuals. Note that the performance has degraded when compared to the Feb 2006 and Nov 2006 cases. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point near by but outside of the domain) is up to 10 cm lower than the observed residual at Neah Bay, WA (a point near by but inside of the domain). Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the cause.
